A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-11-21 analysis/paper/paper.Rmd
Set up plotting. Chekc displacement curve for Løvås sites (cf. negative values)

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial. Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012; @lewis2021], a similar approach is adopted here.

@@

Relative sea-level can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015\]. Variation in this relative distance follow from a range of effects \[e.g. @mörner1976; @milne2009]. Of central importance here is eustasy and glacial istostasy. The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015\]. The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that has volumetric impact on ocean basins. Istostasy, on the other hand, pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight \[@milne2015].

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift has been more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g. @svendsen1987; @romundset2015\]. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle. Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas \[@bjerck2008]. Comparatively, given a continuous and relatively rapid shoreline regression from as high as c. 210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases. In sum, this makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.

@@

Based on the premise that most Stone Age sites located lower than the marine limit were situated on or close to the shoreline, the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island. In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft. This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

@@

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age. Spatial data in the form of features and site limits as defined by the excavating archaeologists was retrieved from local databases at the Museum of Cultural History. Each radiocarbon date has been associated with site features or where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

@@

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

@@

As the excavation of archaeological sites typically follow from residential and commercial development, and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the rgrass7 package. In addition to code and original spatial data being available in the research compendia for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

@@

>>>>

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently spatial. Taking inspiration from recent studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @crema2010; @crema2012], a similar approach is adopted here.

@@

Relative sea-level can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015\]. Variation in this relative distance follow from a range of effects \[e.g. @mörner1976; @milne2009]. Of central importance here is eustasy and glacial istostasy. The eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins. Istostasy, on the other hand, pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet, the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift has been more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable sea-level, been subject to past marine transgression phases, or are still inundated today [e.g. @svendsen1987; @romundset2015]. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle. Transgression, on the other hand, can lead to complete erosion of the sites, or bury them in marine sediments, leading for example to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 210 m above present sea level in the inner Oslo fjord , any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgression phases. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over time, and for identifying potential deviations from this pattern.

@@

Previous evaluations of the correspondence between dates informed by radiocarbon dates and shoreline dating have typically involved a visual inspection of radiocarbon probability distributions plotted against local shoreline displacement curves based on the site elevation [@breivik2018; @solheim2020; @åstveit2018\]. This approach has a couple of limitations. Firstly, only a couple of studies have taken the variable uplift-rates into account when performing this comparison . Secondly, the question of interest in this case is effectively one of synchroneity between events, as opposed to difference \[@parnell2008]. This framing consequently flips the direction of the common null-hypothesis of no significant difference too one of equivalence. While two overlapping confidence intervals means that one cannot conclude that they are different, this does not necessarily mean that they are the same. One implication of this method is consequently that the wider the uncertainty range for either radiocarbon- or shoreline date, the higher the liklihood that the confidence intervals overlap, and the higher the likelihood that we conclude in favour of our hypothesis. The final limitation of this method is that it only takes the vertical distance between site and shoreline into account. While this is the parameter of interest for shoreline dating, the practical implications of a vertical difference in elevation will be highly dependent on the local topography.

In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, or if would have been located on an island. In these cases it would make intuitive sense that the location was used when the ocean reached the elevation of the site, which would then have been accessible by means of watercraft. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

Based on the generally accepted premise that most Stone Age sites located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shorebound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which can be guided by a mental or digital reconstruction of the landscape. typically do not produce material suitable for radiocarbon dating. the most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this.

@@

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age. Spatial data in the form of features and site limits as defined by the excavating archaeologists was retrieved from local databases at the Museum of Cultural History. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

@@

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. To find the adjusted displacement curve for a given site location, the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

@@

As the excavation of archaeological sites typically follow from residential and commercial development, and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the package rgrass7. In addition to code and original spatial data being available in the research compendium for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

@@

Some sources of likely varation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, have not been taken into account. Secondly, the DTM has a vertical error of. Thirdly, the tidal range has not been considered. This is around 0.5 m within the study area today, and there is no reason to think this was substantially different in the past. Fourthly, the exposure of the sites to wave action is also likely to have been of concern, and will presumably have implications for how close to the shoreline people settled.
